---
title: "Effect of absenteeism on graduation rates"
author: "Helen Levy-Myers and Matthew Townley"
date: "2/11/2019"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r gbm_setup, echo = F, include = F, eval = T}

libs = c("xgboost", "caret", "rsample", "pdp", "lime")
# lapply(libs, install.packages, character.only = T)
lapply(libs, library, character.only = T)

# read our data
a_frame = read.csv(file = "y1314_clean.csv", as.is = T)
# First make sure we have rates only $0 <= r <=1$
dim(a_frame)
a_frame = a_frame[which(!is.na(a_frame$abs_r)),]
a_frame = a_frame[which(a_frame$abs_r <= 1 & a_frame$abs_r > 0),]
# a_frame$abs_r = round(a_frame$abs_r, 2)

a_frame = a_frame[which(!is.na(a_frame$grad_r)),]
a_frame = a_frame[which(a_frame$grad_r <= 1 & a_frame$grad_r > 0),]
# a_frame$grad_r = round(a_frame$grad_r, 2)
dim(a_frame)

set.seed(1274)
g_split <- initial_split(a_frame, prop = (3/4))
g_train <- training(g_split)
g_test <- testing(g_split)


features_train = g_train[,-1]
response_train = g_train[,"grad_r"]

features_test = g_test[,-1]
response_test = g_test[,"grad_r"]

```

# 

This is an ecological analysis of 20,000 high schools in the united states to understand whether high rates of chronic absenteeism predict low graduation rates.

# Data

These data come from ....

<https://www2.ed.gov/about/inits/ed/edfacts/data-files/index.html#acgr>
<https://ocrdata.ed.gov/DataFileUsersManual>


# Measures

- Graduation rate
  - Number of graduating students divided by
  - The number of students who started (9th grade?) net of in/outmigration

- Absenteeism rate
  - ? proportion of students who have missed (5? 15? days)
  - 

- Covariates

  - Structural conditions (teacher certification, teacher absenteeism)
  - administrative care (civil rights, sexual assault, violations, disciplinary actions)
  - student engagement/achievement (sports participation, enrollment in higher math, AP)

# Question

Will reducing absenteeism improve high school graduation rates? And if so, can the analysis suggest any approaches to reducing absenteeism?

# Data

```{r eda_plots, echo = F, include = F}

require(hexbin)
hbin = hexbin(a_frame$abs_r, a_frame$grad_r, xbins = 20, xlab = "Absenteeism Rate", ylab = "Graduation Rate")
plot(hbin)
# with(a_frame, lines(abs_r ~ grad_r))
# abline(lm(grad_r ~ abs_r, data = a_frame))
```

```{r}
cor(a_frame) %>% round(2)
:w
```

# Relationship between absenteeism and graduation

# Causal model



